% Conclusions and discussion Chapter

\begin{itemize}
\item CONCLUSIONS: reference to purpose of study
\item CONCLUSIONS: value of / reasons for the study
\item CONCLUSIONS: review of important findings / conclusions
\item CONCLUSIONS: comments, explanations or speculations about findings
\item CONCLUSIONS: limitations of study
\item CONCLUSIONS: implications of study or generalizations
\item CONCLUSIONS: recommendations for future or practical applications - USUALLY SKIPPED
\end{itemize}

This thesis took a look what network based threats could face
installation infrastructure and then studied what kind of means could
be used to protect the initial phases (before OS kernel took the
control of excecution) of installation process using encryption and
file signing.

Protecting every step of communications over networks is important and
protecting installation infrastructure is no exception. This thesis
has shown that it's possible to take a step further in a more secure
installation infrastructure by using two technologies: encryption and
file signatures.

More secure systems can be build step by step by combining simple
individual components without the need for designing a whole new
systems and techonologies. Replacing old components (like TFTP, NFS
and HTTP protocols) with new ones (HTTPS protocol) and increasing the
use of file signatures is a small steps to take for big benefits in
security.

Linux distributions and other open source operating systems use GPG or
other file signing methods to protect the installation packages from
outside tampering which is a really good and important thing to
do. Some Linux distributions also protect the package database
metadata with file signatures, but some distributions have that
functionality turned off by default. Maybe mirrors at some point could
take step forward and enable HTTPS so files like kernel and initrd,
and package database metadata could be securely downloaded?

The public key to verify file signatures is also embedded into initrd
file. Is the initrd file downloaded and verified so that the embedded
public key can be trusted by the installation process?

More testing and verification should be performed for iPXE and it's
TLS implementation and file signing capabilities. This was
intentionally left out from this thesis.
